---
title: "EDA on house prices"
author: "Sanjana Suresh"
date: "18/Sep/2023"
output:
  pdf_document: default
  html_document: default
---
## **Sanjana Suresh **

### **Dataset:** Exploratory Data Analysis on factors affecting housing prices

### **About:** 
### Size of dataset: 528 KB  
### Features:
### The dataset consists of the attributes such as 
### 1. Number of bedrooms, bathrooms, floors
### 2. Condition - Rated in a range from 1 to 5
### 2. Square feet of the house
### 3. Year built
### 4. Location
### 5. Price of the property, etc..
### All of the attributes are real in this data

### Step 1: Load & View the Data

```{r}
#Loading house prices csv file
options(repos = "https://cran.rstudio.com/") 
housing_csv<-read.csv("/Users/sanjana/Desktop/MS/UTA/PS/Dataset/housePricing.csv")


#Displaying first 6 rows of the dataset
head(housing_csv)
#Getting the dimensions of the dataset
dim(housing_csv) 
```
This shows that there are 4600 rows and 18 columns in the given dataset

### Step 2: Summarize the Data

Lets first start our analysis by getting a summary using describe()

**describe()** - It is a part of the Hmisc library. Provides critical statistical information about the dataset such as missing, distinct values,mean, lowest and highest values,etc.. 
```{r message=FALSE}
#Importing the Hmisc library that contains the describe()
library(Hmisc)
#Applying describe() to the dataset
print(describe(housing_csv)) 
```
From the above we can understand that there are no missing values and hence we can move to the next step in the data wrangling process 

## Step 3: Visualize the Data using ggplot

1. Histogram depicting price distribution

```{r}
library(ggplot2)
library(scales)
#Setting up the aesthetic of the plot by mapping price along x axis
ggplot(data=housing_csv, aes(x=price)) +
  
#Adding a histogram layer to the ggplot object and setting interval width a 1000  
geom_histogram(color="brown",binwidth=700000) +
scale_x_continuous(breaks = seq(0, max(housing_csv$price), by = 5000000),labels = comma_format(scale = 1e-3) ) +   
ggtitle("Graph 1 : price vs count")
```


Here,

- scale_x_continuous is used to avoid exponential values and get a scaled range.

- breaks argument creates breaks at intervals of 5000000 starting from 0 up to the maximum value of the "price" variable  

- labels argument is used to format labels with commas and scale values down by a factor of 1000 to make them more readable. 

From the histogram above, we understand the following:

 1. Maximum number of houses have an average price of around $600,000 
 
 2. Most houses are within $5000000
 
 3. Count and price are inversely proportional, i.e the number of houses keeps decreasing as the price of the house gets expensive
 
2. Scatter plot of impact of sq feet on price based on condition

```{r}
#Using the log function so as to distribute the data and compress due to high volume

p <- ggplot(housing_csv, aes(log(sqft_living), log(price), shape = factor(condition)))
p +
  geom_point(aes(colour = factor(condition)), size =9,alpha=0.8) +
   scale_x_continuous(labels = comma) +
   scale_y_continuous(labels = comma) +
ggtitle("Graph 2: sqft vs price wrt condition of house")  
```

From the scatterplot we understand that:

- There are more number of houses with a mediocre condition of 3 and the price is also average. 
  
- There is only one house that is of poor condition 
  
- The price range of houses of very good quality range between 11.25 and 15. However, There is one house that is very expensive and square feet is also less.

- Strong, negative relationship: For an optimal solution, as the variable on the x-axis increases, the variable on the y-axis should decrease. The dots are packed tightly together, which indicates a strong relationship. The above depicts a strong relationship as desired.

3. Histogram
```{r}
ggplot(data = housing_csv, aes(x = floors,color=floors)) +
  geom_histogram(colour="white",fill="darkgreen",binwidth = 0.5) +
  labs(title = "Histogram of Floors",
       x = "floors",
       y = "Count")
```

From the histogram, we can understand that there is more availability of 1 and 2 floor houses

4. CORRELATION MATRIX
 
- The correlation matrix is a very useful metric that helps to establish a relationship between two or more variables in a dataset

- The coefficient indicates both the strength of the relationship as well as the direction(Positive or Negative correlation)

- In our case, it helps to determine the factors that strongly and weakly impact the price of a house


```{r}
#Loading the corrplot library
library(corrplot)
#Choosing the numeric columns for which correlation is to be calculated
num_housing <- housing_csv[,c("price","bedrooms","sqft_living","floors","sqft_lot", "condition", "view", "yr_built", "yr_renovated")]
#Building the correlation matrix
cor_housing <- cor(num_housing)
cor_housing
#Using the corrplot to develop a correlation plot to get a better visualisation of the relationship between variables
corrplot(cor_housing, tl.col = "brown", tl.srt = 30, bg = "White",
         mar = c(3,3,3,3),title = "\n\n Correlation Plot Of Housing Prices Data",type = "full")
```

From the correlation plot, we derive the following:

- There is a negative relationship(-0.3213) of year built with the year renovated i.e properties that were built more recently tend to be less likely to have been renovated.

- The correlation between "price" and "view" is moderately positive (0.2285), indicating that there is a positive relationship between the view of the property and its price. In other words, properties with better views tend to have higher prices.

- "sqft_living" also has a moderate positive correlation with "price" (0.3110), suggesting that there is a positive relationship between the size of the living space and the price of the property. Larger living spaces tend to have higher prices.

- A coefficient of 0.46748066 between "year built" and the "number of floors" of a property suggests a moderate positive linear relationship between these two variables i.e recent properties have more number of floors.

### Step 4: Identify Missing Values

```{r}
sapply(housing_csv, function(x) sum(is.na(x)))
```


- The result indicates that there are no missing values in any of the factors impacting housing prices